---
title: "Beta_Diversity_Final"
author: "denali"
date: "2026-04-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load all necessary libraries
```{r}
library(phyloseq)
library(tidyverse)
library(vegan)
library(dplyr)
library(ggplot2)
```

Data wrangling (selecting only PD patients and binning antioxidants int "High", "Typical", and "Low" intake groups)

```{r}
#load the metadata
metadata = read.delim('project_data/parkinsons_metadata_2.txt', header = TRUE, row.names = 1)

#select only PD patients
metadata_PD <- subset(metadata, Disease == "PD")

#find 33rd and 67th percentiles for the intake of the 8 dietary antioxidants for males and females
PD_sum = metadata_PD |>
  group_by(Sex) |>
  summarise(
    p33_alpha_carotene = quantile(Alpha_carotene, 0.33, na.rm = TRUE),
    p67_alpha_carotene = quantile(Alpha_carotene, 0.67, na.rm = TRUE),
    p33_beta_carotene = quantile(Beta_carotene, 0.33, na.rm = TRUE),
    p67_beta_carotene = quantile(Beta_carotene, 0.67, na.rm = TRUE),
    p33_Copper = quantile(Copper, 0.33, na.rm = TRUE),
    p67_Copper = quantile(Copper, 0.67, na.rm = TRUE),
    p33_selenium = quantile(Selenium, 0.33, na.rm = TRUE),
    p67_selenium = quantile(Selenium, 0.67, na.rm = TRUE),
    p33_Vitamin_A = quantile(Vitamin_A, 0.33, na.rm = TRUE),
    p67_Vitamin_A = quantile(Vitamin_A, 0.67, na.rm = TRUE),
    p33_Vitamin_C = quantile(Vitamin_C, 0.33, na.rm = TRUE),
    p67_Vitamin_C = quantile(Vitamin_C, 0.67, na.rm = TRUE),
    p33_Vitamin_E = quantile(Vitamin_E, 0.33, na.rm = TRUE),
    p67_Vitamin_E = quantile(Vitamin_E, 0.67, na.rm = TRUE),
    p33_Car_total = quantile(Carotene_total, 0.33, na.rm = TRUE),
    p67_Car_total = quantile(Carotene_total, 0.67, na.rm = TRUE))

#bin age of onset
PD_grouped = metadata_PD %>% 
  mutate(Age_onset =
           case_when(Age_onset  < 50 ~ "40s",
                     between(Age_onset, 50, 59) ~ "50s",
                     between(Age_onset, 60, 69) ~ "60s",
                     Age_onset >= 70 ~ "70s"))


#bin the 8 antioxidants according to the values obtained earlier
PD_binned = PD_grouped |>
  mutate(Alpha_carotene =
           case_when(Sex == "Male" & Alpha_carotene > 548.2559 ~ "High",
                     Sex == "Male" & Alpha_carotene < 189.350 ~ "Low",
                     Sex == "Male" & Alpha_carotene ~ "Typical",
                     Sex == "Female" & Alpha_carotene > 575.7513 ~ "High",
                     Sex == "Female" & Alpha_carotene < 541.904 ~ "Low",
                     Sex == "Female" & Alpha_carotene ~ "Typical"),
         Beta_carotene = 
           case_when(Sex == "Male" & Beta_carotene > 3542.550 ~ "High",
                     Sex == "Male" & Beta_carotene < 2229.975 ~ "Low",
                     Sex == "Male" & Beta_carotene ~ "Typical",
                     Sex == "Female" & Beta_carotene > 5100.328 ~ "High",
                     Sex == "Female" & Beta_carotene < 3178.988 ~ "Low",
                     Sex == "Female" & Beta_carotene ~ "Typical"),
         Copper = 
           case_when(Sex == "Male" & Copper > 1.2 ~ "High",
                     Sex == "Male" & Copper < 0.8431575 ~ "Low",
                     Sex == "Male" & Copper ~ "Typical",
                     Sex == "Female" & Copper > 1.285575 ~ "High",
                     Sex == "Female" & Copper < 1 ~ "Low",
                     Sex == "Female" & Copper ~ "Typical"),
         Selenium = 
           case_when(Sex == "Male" & Selenium > 58.39314 ~ "High",
                     Sex == "Male" & Selenium < 42.72238 ~ "Low",
                     Sex == "Male" & Selenium ~ "Typical",
                     Sex == "Female" & Selenium > 65.45464 ~ "High",
                     Sex == "Female" & Selenium < 44.83593 ~ "Low",
                     Sex == "Female" & Selenium ~ "Typical"),
         Vitamin_A =
           case_when(Sex == "Male" & Vitamin_A > 328.4200 ~ "High",
                     Sex == "Male" & Vitamin_A < 191.623 ~ "Low",
                     Sex == "Male" & Vitamin_A ~ "Typical",
                     Sex == "Female" & Vitamin_A > 419.0756 ~ "High",
                     Sex == "Female" & Vitamin_A < 223.012 ~ "Low",
                     Sex == "Female" & Vitamin_A ~ "Typical"),
         Vitamin_C =
           case_when(Sex == "Male" & Vitamin_C > 119.2655 ~ "High",
                     Sex == "Male" & Vitamin_C < 82.31294 ~ "Low",
                     Sex == "Male" & Vitamin_C ~ "Typical",
                     Sex == "Female" & Vitamin_C > 165.2305 ~ "High",
                     Sex == "Female" & Vitamin_C < 108.04810 ~ "Low",
                     Sex == "Female" & Vitamin_C ~ "Typical"),
         Vitamin_E =
           case_when(Sex == "Male" & Vitamin_E > 12.09911 ~ "High",
                     Sex == "Male" & Vitamin_E < 8.652149 ~ "Low",
                     Sex == "Male" & Vitamin_E ~ "Typical",
                     Sex == "Female" & Vitamin_E > 14.34675 ~ "High",
                     Sex == "Female" & Vitamin_E < 10.580000 ~ "Low",
                     Sex == "Female" & Vitamin_E ~ "Typical"),
         Carotene_total =
           case_when(Sex == "Male" & Carotene_total > 4064.849 ~ "High",
                     Sex == "Male" & Carotene_total < 2495.81 ~ "Low",
                     Sex == "Male" & Carotene_total ~ "Typical",
                     Sex == "Female" & Carotene_total > 5691.588 ~ "High",
                     Sex == "Female" & Carotene_total < 3636.83 ~ "Low",
                     Sex == "Female" & Carotene_total ~ "Typical"))
```

Make phyloseq object for the binned data

```{r}
#load data for taxonomy, counts and tree
taxonomy = read.delim('project_data/taxonomy.tsv', row.names = 1) # from qiime2
counts = read.delim('project_data/feature-table.txt', skip=1, row.names=1)  # from qiime2
tree = read_tree("project_data/tree.nwk")


#split Taxon column into multiple columns
taxonomy_formatted = taxonomy %>%
  separate(col = Taxon,
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
           sep = ";", fill = "right") %>%
  select(-Confidence) %>%
  as.matrix()

counts_formatted = counts %>%
  as.matrix()

#create phyloseq object
ps = phyloseq(sample_data(PD_binned),
              otu_table(counts_formatted, taxa_are_rows = TRUE),
              tax_table(taxonomy_formatted), tree)

#save phyloseq object
saveRDS(ps, "phyloseq_taxonomy_binned.rds")
```

Bray curtis beta diversity for the dietary antioxidants

```{r}

set.seed(421)

#rarify data
psrare = ps %>% rarefy_even_depth(sample.size = 9555, rngseed = 421) # 46 samples removed for being lower than sample size

##Alpha Carotene##

#only include samples where alpha carotene values are not missing (NA)
ps_alpha <- subset_samples(psrare, !is.na(Alpha_carotene))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_alpha, method = "bray")

#extract sample metadata
alpha_metadata <- sample_data(ps_alpha) %>% data.frame()

#perform PERMANOVA
Alpha_stats_univar = adonis2(ps_bray ~ Alpha_carotene, data = alpha_metadata)

#combine both into a table
Alpha_stats = bind_rows('Univariate' = Alpha_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 

#rename Pr(>F) and remove missing p-values
Alpha_stats = Alpha_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Alpha_stats 


##Beta Carotene##

#only include samples where beta carotene values are not missing (NA)
ps_beta <- subset_samples(psrare, !is.na(Beta_carotene))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_beta, method = "bray")

#extract sample metadata
beta_metadata <- sample_data(ps_beta) %>% data.frame()

#perform PERMANOVA
Beta_stats_univar = adonis2(ps_bray ~ Beta_carotene, data = beta_metadata)

#combine both into a table
Beta_stats = bind_rows('Univariate' = Beta_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model')

#rename Pr(>F) and remove missing p-values
Beta_stats = Beta_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Beta_stats 

##Total Carotene##

#only include samples where total carotene values are not missing (NA)
ps_total <- subset_samples(psrare, !is.na(Carotene_total))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_total, method = "bray")

#extract sample metadata
total_metadata <- sample_data(ps_total) %>% data.frame()

#perform PERMANOVA
Total_stats_univar = adonis2(ps_bray ~ Carotene_total, data = total_metadata)

#combine both into a table
Total_stats = bind_rows('Univariate' = Total_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model')

#rename Pr(>F) and remove missing p-values
Total_stats = Total_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Total_stats


##Copper##

#only include samples where copper values are not missing (NA)
ps_copper <- subset_samples(psrare, !is.na(Copper))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_copper, method = "bray")

#extract sample metadata
copper_metadata <- sample_data(ps_copper) %>% data.frame()

#perform PERMANOVA
Copper_stats_univar = adonis2(ps_bray ~ Copper, data = copper_metadata)

#combine both into a table
Copper_stats = bind_rows('Univariate' = Copper_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 

#rename Pr(>F) and remove missing p-values
Copper_stats = Copper_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Copper_stats


##Selenium##

#only include samples where selenium values are not missing (NA)
ps_selenium <- subset_samples(psrare, !is.na(Selenium))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_selenium, method = "bray")

#extract sample metadata
selenium_metadata <- sample_data(ps_selenium) %>% data.frame()

#perform PERMANOVA
Selenium_stats_univar = adonis2(ps_bray ~ Selenium, data = selenium_metadata)

#combine both into a table
Selenium_stats = bind_rows('Univariate' = Selenium_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 

#rename Pr(>F) and remove missing p-values
Selenium_stats = Selenium_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Selenium_stats 


##Vitamin A##

#only include samples where vitamin C values are not missing (NA)
ps_vitA <- subset_samples(psrare, !is.na(Vitamin_A))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_vitA, method = "bray")

#extract sample metadata
vitA_metadata <- sample_data(ps_vitA) %>% data.frame()

#perform PERMANOVA
Vit_A_stats_univar = adonis2(ps_bray ~ Vitamin_A, data = vitA_metadata)

#combine both into a table
Vit_A_stats = bind_rows('Univariate' = Vit_A_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 

#rename Pr(>F) and remove missing p-values
Vit_A_stats = Vit_A_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Vit_A_stats 


##Vitamin E##

#only include samples where vitamin C values are not missing (NA)
ps_vitE <- subset_samples(psrare, !is.na(Vitamin_E))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_vitE, method = "bray")

#extract sample metadata
vitE_metadata <- sample_data(ps_vitE) %>% data.frame()

#perform PERMANOVA
Vit_E_stats_univar = adonis2(ps_bray ~ Vitamin_E, data = vitE_metadata)

#combine both into a table
Vit_E_stats = bind_rows('Univariate' = Vit_E_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 

#rename Pr(>F) and remove missing p-values
Vit_E_stats = Vit_E_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Vit_E_stats 

##Vitamin C##

#only include samples where vitamin C values are not missing (NA)
ps_vitC <- subset_samples(psrare, !is.na(Vitamin_C))

#use Bray-Curtis distance
ps_bray <- phyloseq::distance(ps_vitC, method = "bray")

#extract sample metadata
vitc_metadata <- sample_data(ps_vitC) %>% data.frame()

#perform PERMANOVA
Vit_C_stats_univar <- adonis2(ps_bray ~ Vitamin_C, data = vitc_metadata)

#combine both into a table
Vit_C_stats = bind_rows('Univariate' = Vit_C_stats_univar %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model')

#rename Pr(>F) and remove missing p-values
Vit_C_stats = Vit_C_stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

Vit_C_stats 
```

Place all p and R2 values in a table and calculate the adjusted p-values - then display in a bar plot
```{r}

#make one table coombining the pvalues and R2 values for all antioxidants

results <- tibble(
  Variable = c("Alpha_carotene","Beta_carotene","Carotene_total","Copper",
               "Selenium","Vitamin_A","Vitamin_C","Vitamin_E"),
  Pval = c(
    Alpha_stats$Pval,
    Beta_stats$Pval,
    Total_stats$Pval,
    Copper_stats$Pval,
    Selenium_stats$Pval,
    Vit_A_stats$Pval,
    Vit_C_stats$Pval,
    Vit_E_stats$Pval
  ),
  R2 = c(Alpha_stats$R2,
    Beta_stats$R2,
    Total_stats$R2,
    Copper_stats$R2,
    Selenium_stats$R2,
    Vit_A_stats$R2,
    Vit_C_stats$R2,
    Vit_E_stats$R2
  )
)

#calculate the adjusted pvalues using false discovery rate
results$Pval_adj <- p.adjust(results$Pval, method = "fdr")

#plot R2 values in a bar plot to identify the antioxidant that is responsible for most beta diversity variation
results_pvalue <- ggplot(data = results, aes(x = reorder(Variable, R2, decreasing = FALSE),y=R2, fill = Variable)) +
  geom_bar(stat='identity') +
  coord_flip() + 
  labs(y = "Adonis R2", x = "Antioxidant", title = "Antioxidants Beta Diversity R2 Results")+
  theme_classic(base_size = 15)+
  theme(legend.position = "none")

results_pvalue
```

Make the Non-Metric Multidimensional Scaling (NMDS) plot for Vitamin C

```{r}
#perform NMDS
mds = metaMDS(ps_bray)

#combine NMDS results with vitamin C metadata
mds_data_vitC = mds$points %>% as.data.frame %>%
  merge(sample_data(ps_vitC), by='row.names', sort=F)
head(mds_data_vitC)

#reorganize vitamin C intake groups for low, typical, and high in order
mds_data_vitC$Vitamin_C <- factor(mds_data_vitC$Vitamin_C,
                           levels = c("Low", "Typical", "High"))

#produce vitamin C NMDS plot
p_vitC = mds_data_vitC %>%
  ggplot(aes(MDS1,MDS2,color = Vitamin_C, shape = Vitamin_C)) +
  geom_point() +
  stat_ellipse() +
  labs(title= str_wrap('Bray-Curtis Beta Diversity Across Vitamin C Intake Groups', width = 50)) + 
  scale_color_manual(name = "Vitamin C Intake", values = c("High" = "#F5793A", "Low" = "#E03C8A", "Typical" = "#005AB5")) +
  scale_shape_manual(name = "Vitamin C Intake", values = c("High" = 16, "Low" = 17, "Typical" = 15))+
  theme_classic(base_size=15) +
  theme(plot.title = element_text(hjust = 0.5))
p_vitC

ggsave("beta_diversity_vitC.png",
       p_vitC,
       width = 8,
       height = 5,
       dpi = 300)
```


