---
title: "assignment6"
output: html_document
date: "2026-02-28"
---

## Key information 

Dataset: Parkinson's Disease

Group of interest: Vitamin C intake levels (low vs high)

Filtering:

1) Keep only PD patients (Disease == 'PD')

2) Filter out individuals with 'Typical' Vitamin C levels (keep 'Low' and 'High'). This categorization was done in previous assignments. 

## Loading packages

```{r}
library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(Maaslin2)
library(GGally)
library(vegan)
```

# Vitamin C
## Importing files

```{r}
# Phyloseq object from previous assignments

meta = readRDS('phyloseq.rds') %>% 
  .@sam_data %>% 
  data.frame() %>% 
  rownames_to_column('sample_name')
path_abun = read.delim('path_abun_unstrat.tsv', row.names = 1)
```

## Filtering

Phyloseq object imported above only contains patiens with PD. Controls are already filtered out. 

The code below filters for extreme Vitamin C intake levels (Low and High). 
```{r}
meta_vitaminc = meta %>% 
  filter(Vitamin_C %in% c("Low", "High"))
path_abun_vitaminc = path_abun %>% 
  select(all_of
         (meta_vitaminc$sample_name))
```

## Differential Abundance Analysis using MaAsLin2 method

```{r}
daa_results_df = pathway_daa(abundance = path_abun_vitaminc,
                             metadata = meta_vitaminc, 
                             group = "Vitamin_C", 
                             daa_method = "LinDA", 
                             select = NULL, reference = NULL)

daa_annotated_results_df = pathway_annotation(pathway = "MetaCyc",
                                              daa_results_df = daa_results_df,
                                              ko_to_kegg = FALSE) %>% 
  rename(coeff = log2_fold_change)
saveRDS(daa_annotated_results_df,'vitaminc_diff_abund_results_df.rds')

```

## Calculating Log 2 Fold Change 

```{r}
pathway_l2fc = path_abun_vitaminc %>% t() %>%
  as.data.frame() %>%
  mutate(Group = meta_vitaminc$Vitamin_C) %>%
  group_by(Group) %>%
  summarise(across(everything(), mean)) %>%
  t() %>% 
  as.data.frame() %>% 
  setNames(.[1,]) %>% 
  slice(-1) %>%
  mutate(across(everything(), as.numeric)) %>% 
  mutate(feature = rownames(.)) %>% 
  mutate(log2_fold_change = log2(High/Low))

daa_annotated_l2fc_df = daa_annotated_results_df %>% 
  left_join(pathway_l2fc, by = 'feature')

head(daa_annotated_l2fc_df)
```

## Filtering for pathways:
- with adjusted P value of <0.05;
- log 2 fold change cutoff of >=1
```{r}
daa_annotated_l2fc_df %>% 
  filter(p_adjust < 0.05) %>% 
  filter(abs(log2_fold_change) >= 1)
```

**Remove**
There are no pathways that fit the above requirement. 
There are no significant pathways.

```{r}
new_pthreshold = daa_annotated_l2fc_df %>%
  arrange(p_adjust) %>% 
  slice(3) %>% 
  pull(p_adjust)

print(paste("In order to have at least 3 significant pathways, adjusted P value threshold has to be moved to", new_pthreshold, '.'))
```
Filtering using the new p value threshold.
**

```{r}
daa_annotated_filtered_df = daa_annotated_l2fc_df %>% 
  filter(p_adjust < 0.05) %>% 
  filter(abs(log2_fold_change) >= 1)  

daa_annotated_filtered_df$description[daa_annotated_filtered_df$feature == "PWY-7873"] = "D-erythronate degradation II"

daa_annotated_filtered_df$description[daa_annotated_filtered_df$feature == "PWY-7874"] = "L-threonate degradation"

daa_annotated_filtered_df %>% 
  select(feature, p_adjust, log2_fold_change) %>% 
  print()
```

## Generating error bar plot

Note 1: I did not know how to change font size using an arguement in pathway_errorbar_fixed function. So I changed it directly in the source r code.

Note 2: The figure below is formatted to publication-ready standards based on png file via ggsave(). It is included in the zipped folder as a standalone png . HTML rendering does not fully show the adjustments.
```{r}
source('ggpicrust2_errorbar_function_fixed.R')

peb = pathway_errorbar_fixed(abundance = path_abun_vitaminc, 
                     daa_results_df = daa_annotated_filtered_df, 
                     Group = meta_vitaminc$Vitamin_C, 
                     wrap_label = T, wraplength=30,
                     fc_cutoff = 1, order_by_log = T,
                     p_values_threshold = 0.05, 
                     order = "group", 
                     ko_to_kegg = FALSE, 
                     p_value_bar = TRUE, 
                     x_lab = "description", 
                     colors = c('#F5793A', '#005AB5'))
ggsave("project_data/vitaminc_daa_error_bar.png",
       plot = peb, width = 18, height = 10)
```

```{r}
pca_vc = pathway_pca(abundance = path_abun_vitaminc,
            metadata = meta_vitaminc,
            group = 'Vitamin_C', 
            colors = c('#F5793A', '#005AB5'))
volcano = pathway_volcano(daa_annotated_l2fc_df,
                       fc_col = "log2_fold_change",
                       p_col = "p_adjust",
                       label_col = ",description",
                       label_top_n = 0,
                       point_alpha = 0.4,
                       colors = c('#F5793A', '#005AB5', '#E03C8A'))
ggsave("project_data/pathway_pca_vc.png",
       plot = pca_vc, width = 8, height = 6)
ggsave("project_data/pathway_volcano.png",
       plot = volcano, width = 9, height = 5)
```

```{r}
dist_matrix <- vegdist(t(path_abun_vitaminc), method = "bray")

permanova_result <- adonis2(dist_matrix ~ Vitamin_C, 
                            data = meta_vitaminc, 
                            permutations = 9999)

print(permanova_result)
```
